Weak interspecific interactions in a sagebrush steppe\new{? Conflicting evidence from observations and experiments}

### 2018-01-27 removals/manuscript/removal_ms_SPE.tex
comments on ms

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% 
%%% Just for commenting
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\usepackage[usenames]{color}
\newcommand{\new}{\textcolor{red}}
\newcommand{\comment}{\textcolor{black}}

@@

Stable coexistence requires intraspecific limitations to be stronger than interspecific limitations. The greater the difference between intra- and interspecific limitations, the more stable the coexistence, and the weaker the competitive release any species should experience following removal of competitors. We conducted a removal experiment to test whether a previously estimated model, showing surprisingly weak interspecific competition for four dominant species in a sagebrush steppe, accurately predicts competitive release. Our treatments were 1) removal of all perennial grasses and 2) removal of the dominant shrub, \textit{Artemisia tripartita}. We regressed survival, growth and recruitment on the locations, sizes, and species identities of neighboring plants, along with an indicator variable for removal treatment. If our ``baseline" regression model, which accounts for local plant-plant interactions, accurately explains the observed responses to removals, then the removal coefficient should be non-significant. \new{For survival, the removal coefficients were never significantly different from zero, and only \textit{A. tripartita} showed a (negative) response to removals at the recruitment stage. For growth, the removal treatment effect was significant and positive for two species, \textit{Poa secunda} and \textit{Pseudoroegneria spicata}, indicating that the baseline model underestimated interspecific competition. For all three grass species, population models based on the vital rate regressions that included removal effects projected 1.4 to 3-fold increases in equilibrium population size relative to the baseline model (no removal effects).} However, including information about pretreatment cover of \textit{A. tripartita} at the quadrat or individual scale failed to improve the growth regressions, raising questions about the mechanisms driving the positive response to removal. \new{While the experimental results showed greater competitive release than our baseline models predicted, we will need more tightly controlled experiments focused on underlying mechanisms to conclusively validate or reject our model's predictions.}

@@

The positive removal effects on \textit{P. secunda}  and \textit{P. spicata} growth suggest that models based on observational data may underestimate competitive release. We conducted additional analyses to further explore removal effects on the growth of these two species. First, \new{we asked whether increases in grass growth following removal were greatest in quadrats or neighborhoods where we removed the most \textit{A. tripartita}. We did this by taking residuals from the fitted growth models for plants in removal quadrats and plotting those residuals against the pre-treatment, quadrat-level cover of \textit{A. tripartita}. No species showed the positive relationship we expected to see (Fig. \ref{fig:GrowthResidsQuad}). In fact, \textit{Poa secunda} residuals were significantly lower in quadrats with high pre-treatment cover of \textit{A. tripartita} (individuals in high cover quadrats are growing more slowly than the growth model predicted). We repeated this analysis at the individual plant level for grasses present in 2011. We calculated pre-treatment crowding from \textit{A. tripartita} for each of those individuals, then plotted their growth residuals against crowding for all years in which each plant survived. We found no significant relationships between the growth residuals and pre-treatment crowding from \textit{A. tripartita} (Fig. \ref{fig:GrowthResidsIndivid}). }

@@

Our population model based on historical, observational data revealed weak interspecific interactions \citep{adler_coexistence_2010,chu_large_2015}. As a result, the model predicts very little competitive release among the four dominant species in a sagebrush steppe. 
Because the model was based solely on observational
data, we were suspicious of its ability to accurately describe the outcome of interspecific interactions, so we conducted a removal experiment to test it. 
Our approach represents a strong test of a surprising prediction for two reasons. First, the prediction was already published; the experiment 
gave us a very real opportunity to be wrong in public. Second, testing an observationally-based model with an experimental manipulation, 
not just additional observations, further increases the potential to confront the model with a novel situation outside the historical range of variation.
\new{We expected that upon completion of this study, we would have clear evidence for the success or failure of our previous model's predictions. Unfortunately, our results do not provide that clarity. In the sections below, we first summarize the evidence that our experiment revealed greater competitive release than our baseline model predicted, we then explain weaknesses in the experimental results that complicate the interpretation, and we end by discussing general lessons for the study of interspecific interactions.}

@@

\new{Another reason we view the growth results with some suspicion is that increases in growth were no larger in individual neighborhoods or quadrats with higher pre-treatment \textit{A. tripartita} cover. If \textit{A. tripartita} removal is releasing grasses from competition, shouldn't this effect be larger for grasses located in areas with high \textit{A. tripartita} cover? The absence of this relationship (Figs. \ref{fig:GrowthResidsQuad} and \ref{fig:GrowthResidsIndivid}) raises questions about the mechanism by which \textit{A. tripartita} removal promotes grass growth. Perhaps \textit{A. tripartita's} roots extend far enough beyond its canopy to compete for resources with grasses located in interspaces. While this might explain the lack of the relationship at the neighbrhood scale, it seems unlikely to explain the quadrat-level results. } 
An alternative explanation involves pre-treatment differences in composition between control plots and  the \textit{A. tripartita} removal plots, which had higher cover of \textit{A. tripartita} (see 2011 means in Fig. \ref{fig:CoverTrends}) and higher cover of subdominant grasses and shrubs (Fig. \ref{fig:W-by-treatment}). These pre-treatment differences could reflect subtle differences in edaphic conditions which could promote \textit{P. spicata} growth. \new{However, since the 1900's, many quadrats have experienced dramatic changes in \textit{A. tripartita} cover, arguing against viewing \textit{A. tripartita} cover as a sensitive indicator of edaphic conditions. 
A final possibility is that our removal treatment influenced grass growth through some other mechanism, unrelated to competition. Maybe the loss of shrub cover benefits grasses by reducing small mammal herbivory \citep{Orrock2010}? Without a better understanding of the mechanism driving increased grass growth in the shrub removal quadrats, we are hesitant to dramatically adjust our prior understanding gained from models fit to decades of observational data.  }

@@

\new{A variety of approaches are available for studying interspecific interactions among plants, and each has its advantages and disadvantages. Observational datasets have the advantage of being realistic: they reflect all the processes and interactions that drive community dynamics across a range of spatial and temporal scales. Their disadvantage is weak inference. Distinguishing between correlation and causation is challenging, and simulations of a model fit to observational data may rely heavily on extrapolation. Experiments, such as response surface designs (Inouye), offer an alternative approach to studying interactions at the population scale. Their chief advantage is strong inference about causality. But they sacrifice realism: experiments are typically implemented at fine spatial and temporal scales which exclude potentially important processes.  Finally, we can study interactions mechanistically by describing patterns of resource use \citep{Dybzinski2007,Silvertown1999}. Mechanistic studies have the virtue of focusing on directly on the processes of interest, but rarely address more than one limiting resource at a time. }

\new{ Our previous work was based purely on observational data, so we were particularly concerned about the extrapolation issue.
Specifically, our model projects the performance of individual plants growing in the absence of neighbors. Although the observational data set seems to include such situations (Fig. \ref{fig:Wscatters}), if these isolated individual are still interacting with neighbors belowground, our model might underestimate the true intrinsic growth rates. Therefore, we conducted the removal experiment, hoping to gain the advantages offered by multiple, complementary approaches. The positive responses of \textit{Poa secunda} and \textit{P. pseudoroegneria} to removals suggests our observational approach might have underestimated intrinsic growth rates after all. But removal experiments fall into a gray area between the observational and experimental approach. They are an experimental manipulation, but they rely on natural spatial variation in composition. We can't remove a plant from a location where it does not occur. The reliance on natural variation weakens inference, leaving nagging questions about pre-existing differences among quadrats. We could have avoided those concerns by conducting a response surface experiment, planting our study species in randomly selected locations. }

\new{We also learned that partial removals, rather than complete removal treatments, could 
have helped us to distinguish between competition-mediated and density-independent effects of removal treatments (e.g. Fig. \ref{fig:PSSPresids}).
Partial removals, or experimental gradients in densities, would give us inference about the competition coefficients themselves, whereas our 
design lumped all treatment effects into the intercept terms. Complete removals might be adequate for testing the presence of competition, but do not work as well for quantifying per capita competitive effects. 
Furthermore, while our experiment focused on a few interspecific interactions, ultimately it is the difference in the strength of intra- and interspecific interactions which determines the stability of coexistence and the degree to which the dynamics of co-occurring species are coupled. Next time we will also include intraspecific removal treatments.  }

\new{The difficulty we experienced interpreting our results also made us appreciate the value of mechanistic insight. Even if the removal experiment had perfectly validated the predictions of our phenomenological baseline model, we suspect we would still be asking ourselves, how can interspecific interactions really be so weak? The only way to rigorously answer that question is to study the factors--resources, natural enemies, environmental constraints--limiting population growth. }

Although our results point to limitations of model based on long-term observational data, they also give some cause for optimism about population forecasting. \citet{Crone2013} found that matrix projection models for plants often failed to forecast population trends due 
to environmental differences between the data collection and forecast periods. Our population models 
performed quite well, after accounting for year-to-year environmental variation through the inclusion of
random year effects.

@@

\begin{figure}[h]
\centering
\includegraphics[width=1\textwidth]{growth_residuals_vs_quad_removal_cover}
\caption{The relationship between residuals from the fitted growth model and quadrat-level pre-treatment cover of the removal species (all grasses removed in quadrats where we studied \textit{A. tripartita}, and \textit{A. tripartita} removed in the quadrats where we studied the grasses). Linear regressions are shown only when statistically significant. Source file:  \texttt{growth\textbackslash growth\textunderscore residuals\textunderscore by\textunderscore quad.R}.}
\label{fig:GrowthResidsQuad}
\end{figure}

\begin{figure}[h]
\centering
\includegraphics[width=1\textwidth]{growth_residuals_vs_Wremoval}
\caption{The relationship between residuals from the fitted growth model and individual-level pre-treatment crowding exerted by the removal species (all grasses in quadrats where we studied \textit{A. tripartita}, and \textit{A. tripartita} in the quadrats where we studied the grasses). Only individuals present in 2011 in the removal treatment quadrats are included in this analysis. If those individuals survived into later years, we did include their performance in those later years. None of the relationships were statistically significant. Source file:  \texttt{growth\textbackslash growth\textunderscore residuals\textunderscore by\textunderscore individ.R}.}
\label{fig:GrowthResidsIndivid}
\end{figure}

>>>>

\graphicspath{ {figures/} }

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% 
%%% Just for commenting
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\usepackage[usenames]{color}
\newcommand{\new}{\textcolor{red}}
\newcommand{\spe}{\textcolor{blue}}
\newcommand{\comment}{\textcolor{black}}

@@

Stable coexistence requires intraspecific limitations to be stronger than interspecific limitations. The greater the difference between intra- and interspecific limitations, the more stable the coexistence, and the weaker the competitive release any species should experience following removal of competitors. We conducted a removal experiment to test whether a previously estimated model, showing surprisingly weak interspecific competition for four dominant species in a sagebrush steppe, accurately predicts competitive release. Our treatments were 1) removal of all perennial grasses and 2) removal of the dominant shrub, \textit{Artemisia tripartita}. We regressed survival, growth and recruitment on the locations, sizes, and species identities of neighboring plants, along with an indicator variable for removal treatment. If our ``baseline" regression model, which accounts for local plant-plant interactions, accurately explains the observed responses to removals, then the removal coefficient should be non-significant. \new{For survival, the removal coefficients were never significantly different from zero, and only \textit{A. tripartita} showed a (negative) response to removals at the recruitment stage. For growth, the removal treatment effect was significant and positive for two species, \textit{Poa secunda} and \textit{Pseudoroegneria spicata}, indicating that the baseline model underestimated interspecific competition. For all three grass species, population models based on the vital rate regressions that included removal effects projected 1.4 to 3-fold increases in equilibrium population size relative to the baseline model (no removal effects).} 
%However, including information about pretreatment cover of \textit{A. tripartita} at the quadrat or individual scale failed to 
improve the growth regressions
\spe{However, we found no evidence of higher response to removal in quadrats with higher pretreatment cover of \textit{A. tripartita}, or by plants with higher pre-treatment crowding by \textit{A. tripartita} individuals,} 
raising questions about the mechanisms driving the positive response to removal.
\new{While the experimental results showed greater competitive release than our baseline models predicted, we will need more tightly controlled experiments focused on underlying mechanisms to conclusively validate or reject our model's predictions.}

@@

The positive removal effects on \textit{P. secunda}  and \textit{P. spicata} growth suggest that models based on observational data may underestimate competitive release. We conducted additional analyses to further explore \spe{this hypothesis.} 
%removal effects on the growth of these two species. 
First, \new{we asked whether increases in grass growth following removal were greatest in quadrats or neighborhoods where we removed the most \textit{A. tripartita}. We did this by taking residuals from the fitted growth models for plants in removal quadrats and plotting those residuals against the pre-treatment, quadrat-level cover of \textit{A. tripartita}. No species showed the positive relationship 
\spe{that would be expected} 
% we expected to see 
(Fig. \ref{fig:GrowthResidsQuad}). In fact, \textit{Poa secunda} residuals were significantly lower in quadrats with high pre-treatment cover of \textit{A. tripartita} (individuals in high pre-treatment cover quadrats grew more slowly than the growth model 
\spe{including the removal effect} predicted). We repeated this analysis at the individual plant level for grasses present in 2011. We calculated pre-treatment crowding from \textit{A. tripartita} for each of those individuals, then plotted their growth residuals against crowding for all years in which each plant survived. We found no significant relationships between the growth residuals and pre-treatment crowding from \textit{A. tripartita} (Fig. \ref{fig:GrowthResidsIndivid}). }

@@

Our population model based on historical, observational data revealed weak interspecific interactions \citep{adler_coexistence_2010,chu_large_2015}. As a result, the model predicts very little competitive release among the four dominant species in a sagebrush steppe. 
Because the model was based solely on observational
data, we were suspicious of its ability to accurately describe the outcome of interspecific interactions, so we conducted a removal experiment to test it. 
Our approach represents a strong test of a surprising prediction for two reasons. First, the prediction was already published; the experiment 
gave us a very real opportunity to be wrong in public. Second, testing an observationally-based model with an experimental manipulation, 
not just additional observations, further increases the potential to confront the model with a novel situation outside the historical range of variation.
\spe{We expected this study to provide} \new{clear evidence for the success or failure of our previous model's predictions. Unfortunately, our results do not provide that clarity. In the sections below, we first summarize the evidence that our experiment revealed greater competitive release than our baseline model predicted, then explain weaknesses in the experimental results that complicate the interpretation, and end by discussing general lessons for the study of interspecific interactions.}

@@

\new{Another reason we view the growth results with some suspicion is that increases in growth were no larger in individual neighborhoods or quadrats with higher pre-treatment \textit{A. tripartita} cover. If \textit{A. tripartita} removal is releasing grasses from competition, shouldn't this effect be larger for grass genets in areas with high \textit{A. tripartita} cover? The absence of this relationship (Figs. \ref{fig:GrowthResidsQuad} and \ref{fig:GrowthResidsIndivid}) raises questions about the mechanism by which \textit{A. tripartita} removal promotes grass growth. Perhaps \textit{A. tripartita's} roots extend far enough beyond its canopy to compete for resources with grasses located in interspaces. While this might explain the lack of the relationship at the neighborhood scale, it seems unlikely to explain the quadrat-level results.}
An alternative explanation involves pre-treatment differences in composition between control plots and  the \textit{A. tripartita} removal plots, which had higher cover of \textit{A. tripartita} (see 2011 means in Fig. \ref{fig:CoverTrends}) and higher cover of subdominant grasses and shrubs (Fig. \ref{fig:W-by-treatment}). These pre-treatment differences could reflect subtle differences in edaphic conditions which could promote \textit{P. spicata} growth. \new{However, since the 1900's, many quadrats have experienced dramatic changes in \textit{A. tripartita} cover, arguing against viewing \textit{A. tripartita} cover as a sensitive indicator of edaphic conditions. 
A final possibility is that our removal treatment influenced grass growth through some other mechanism, unrelated to competition.} 
\spe{For example, loss of shrub cover might benefits grasses by reducing small mammal herbivory \citep{Orrock2010}. 
Without a better understanding of the mechanism driving increased grass growth in the shrub removal quadrats, 
we are hesitant to dramatically adjust our prior understanding gained from models fit to decades of observational data.  }

@@

\new{A variety of approaches are available for studying interspecific interactions among plants, and each has its advantages and disadvantages. Observational datasets have the advantage of \spe{realism}: they reflect all the processes and interactions that drive community dynamics across a range of spatial and temporal scales. Their disadvantage is weak inference. Distinguishing between correlation and causation is challenging, and simulations of a model fit to observational data may rely heavily on extrapolation. Experiments, such as response surface designs (Inouye), offer an alternative approach to studying interactions at the population scale. Their chief advantage is strong inference about causality. But they sacrifice realism: experiments are typically implemented at fine spatial and temporal scales which exclude potentially important processes.  Finally, we can study interactions mechanistically by describing patterns of resource use \citep{Dybzinski2007,Silvertown1999}. Mechanistic studies have the virtue of focusing on directly on the processes of interest, but rarely address more than one limiting resource at a time. }

\new{Our previous work was based purely on observational data, so we were particularly concerned about the extrapolation issue.
Specifically, our model projects the performance of individual plants growing in the absence of neighbors. Although the observational data set seems to include such situations (Fig. \ref{fig:Wscatters}), if these isolated individual are still interacting with neighbors belowground, our model might underestimate the true intrinsic growth rates \spe{that would occur when near and distant neighbors are removed.} 
\spe{(SPE: I'm not sure if readers will understand the issue here. I think the point you're making is that if our competition kernels
under-estimate the strength of long-range interactions, they will underestimate the effect of removing all ARTR from a quadrat. 
And because they are based on observational data, the estimated kernels might do that. Maybe near-neighbor and far-neighbor density 
are confounded due to quadrat-to-quadrat variation, so it appears that only near neighbors matter. Or maybe because 
near neighbors matter most, the model that optimally balances
realism and parsimony (to optimize prediction) only based predictions on close neighbors.)} 
Therefore, we conducted the removal experiment, hoping to gain the advantages offered by multiple, complementary approaches. The positive responses of \textit{Poa secunda} and \textit{P. pseudoroegneria} to removals suggests our observational approach might have underestimated intrinsic growth rates after all. But removal experiments fall into a gray area between \spe{an observational study and a randomized, controlled experiment}. They are an experimental manipulation, but they rely on natural spatial variation in composition. We can't remove a plant from a location where it does not occur. The reliance on natural variation weakens inference, leaving nagging questions about pre-existing differences among quadrats. We could have avoided those concerns by conducting a response surface experiment, planting our study species in randomly selected locations.}

\new{We also learned that partial removals, rather than complete removal treatments, could 
have helped to distinguish between competition-mediated and \spe{other} effects of removal treatments (e.g. Fig. \ref{fig:PSSPresids}).
Partial removals, or experimental gradients in densities, would give us inference about the competition coefficients themselves, whereas our 
design lumped all treatment effects into the intercept terms. Complete removals might be adequate for testing the presence of competition, but do not work as well for quantifying per capita competitive effects. 
Furthermore, while our experiment focused on a few interspecific interactions, ultimately it is the difference in the strength of intra- and interspecific interactions which determines the stability of coexistence and the degree to which the dynamics of co-occurring species are coupled. Next time we will also include intraspecific removal treatments. }

\new{Our difficulties with interpreting our results also made us appreciate the value of mechanistic insight. Even if the removal experiment had perfectly validated the predictions of our phenomenological baseline model, we suspect we would still be asking ourselves, how can interspecific interactions really be so weak? The only way to rigorously answer that question is to study the factors--resources, natural enemies, environmental constraints--limiting population growth. }

Although our results point to limitations of models based on long-term observational data, they also give some cause for optimism about population forecasting. \citet{Crone2013} found that matrix projection models for plants often failed to forecast population trends due 
to environmental differences between the data collection and forecast periods. 
%Our population models performed quite well after accounting for year-to-year environmental variation through the inclusion of
%random year effects. 
\spe{By including random year-effects to account for year-to-year environmental variation, our population models performed quite well
at forecasting between-year population changes (Figure \ref{fig:obsVpred}). }

@@

\begin{figure}[h]
\centering
%\includegraphics[width=1\textwidth]{growth_residuals_vs_quad_removal_cover}
\caption{The relationship between residuals from the fitted growth model and quadrat-level pre-treatment cover of the removal species (all grasses removed in quadrats where we studied \textit{A. tripartita}, and \textit{A. tripartita} removed in the quadrats where we studied the grasses). Linear regressions are shown only when statistically significant. Source file:  \texttt{growth\textbackslash growth\textunderscore residuals\textunderscore by\textunderscore quad.R}.}
\label{fig:GrowthResidsQuad}
\end{figure}

\begin{figure}[h]
\centering
%\includegraphics[width=1\textwidth]{growth_residuals_vs_Wremoval}
\caption{The relationship between residuals from the fitted growth model and individual-level pre-treatment crowding exerted by the removal species (all grasses in quadrats where we studied \textit{A. tripartita}, and \textit{A. tripartita} in the quadrats where we studied the grasses). Only individuals present in 2011 in the removal treatment quadrats are included in this analysis. If those individuals survived into later years, we did include their performance in those later years. None of the relationships were statistically significant. Source file:  \texttt{growth\textbackslash growth\textunderscore residuals\textunderscore by\textunderscore individ.R}.}
\label{fig:GrowthResidsIndivid}
\end{figure}
